---
title: "Analyzing Iris with R and Python"
author: "Sicong Deng"
date: 2026-09-21
format: html
engine: knitr
---

This analysis demonstrates interoperability between R and Python using
`reticulate`. The Iris dataset is loaded and initially manipulated in R,
passed to Python for further analysis, and then returned to R for
visualization.

## Loading the data in R

```{r}
library(reticulate)
library(readr)
library(dplyr)
library(ggplot2)

iris_data <- read_csv("data/iris.csv")

head(iris_data)
```

The Iris dataset contains measurements of sepal and petal dimensions for
150 flowers belonging to three species. We begin by loading the local
copy of the dataset in R.

## Data manipulation in R

We first calculate the mean petal length and width for each species.

```{r}
species_summary <- iris_data |>
  group_by(Species) |>
  summarise(
    mean_petal_length = mean(Petal.Length),
    mean_petal_width = mean(Petal.Width),
    n = n()
  )

species_summary
```

Grouping the observations by species reduces the original flower-level
dataset to a species-level summary. The resulting `species_summary`
object is an R data frame containing the average petal measurements and
number of observations for each species.

## Passing the data to Python

The R object can be accessed from Python through `r.species_summary`.

```{python}
import pandas as pd

summary_python = r.species_summary

summary_python
```

Python can now use the data produced by the previous R computation. We
calculate the ratio between mean petal length and mean petal width.

```{python}
summary_python["petal_ratio"] = (
    summary_python["mean_petal_length"]
    / summary_python["mean_petal_width"]
)

summary_python
```

The new `petal_ratio` variable is therefore computed in Python from
summary statistics originally calculated in R.

## Passing the result back to R

The Python object is returned to R using `py$summary_python`.

```{r}
final_summary <- py$summary_python

final_summary
```

At this point, `final_summary` contains results produced by both
languages: the group-level statistics were calculated in R, while
`petal_ratio` was calculated in Python.

## Visualization in R

```{r}
#| label: fig-iris-petal-ratio
#| fig-cap: "Ratio of mean petal length to mean petal width by Iris species."

ggplot(
  final_summary,
  aes(
    x = Species,
    y = petal_ratio,
    fill = Species
  )
) +
  geom_col() +
  labs(
    title = "Petal Length-to-Width Ratio by Species",
    x = "Species",
    y = "Mean petal length / mean petal width"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

As shown in @fig-iris-petal-ratio, the ratio between mean petal length
and mean petal width differs among the three Iris species. The values
plotted here depend on computations performed in both R and Python,
demonstrating how objects can be transferred between the two languages
within a single Quarto document.
